---
title: "R Meetup: dplyr overview"
author: "Michael Abla"
date: "July 20, 2019"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: false
---

```{r setup, include=FALSE} 

# library
library(dplyr)
library(magrittr)

# load data
data("USArrests")
force(USArrests)

cities <- read.csv("UScities_simplemaps.csv", stringsAsFactors = F)

```

# agenda

## topics

- overview
- basic functions
- summarizing
- joining
- misc tips

# overview

## what is dplyr?

- introduced by Hadley Wickham in 2014
- derived from `plyr`
- included in `tidyverse` package
- mostly used for data wrangling
- see cheat sheet [here](https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf)

## data used in this presentation
**USArrests** data, available in base R - use `data()`
```{r USAressts}
names(USArrests)
```
**US cities** data from [simplemaps](https://simplemaps.com/data/us-cities)
```{r cities}
names(cities)
```

## a word about %>%
- the package `magrittr` provides a forward-pipe operator `%>%`
- extremely valuable operator for R coding. See example below:

```{r magrittr, echo=TRUE, message=F}
USArrests %>% summary()
```

# basic functions

## popular dplyr verbs

- `mutate()` adds new variables
- `select()` picks variabes based on their names
- `filter()` picks cases based on their values
- `summarise()` reduces multiple values down to a single summary
- `arrange()` changes the ordering of the rows
- `rename()` renames variables
- `sample_n()` and `sample_frac()` to take random samples

## a word about _at, _if, _all

dplyr verbs have three main variations to help with transforming multiple variables:

- `_all` affects every variable
- `_at` affects variables selected (helper function `vars()` commonly used)
- `_if` affects variables selected based on conditions

## mutate example

```{r verb example 1, echo=TRUE,warning=FALSE, message=F}

USArrests %>%
  mutate(AssaultplusMurder = Assault + Murder) %>% # create new var
  mutate_at(vars(UrbanPop, Rape), list(x100 = ~ .*100)) %>% # create multiple vars
  head()
```

## another example of dplry verbs in action

```{r verb example 2, echo=TRUE,warning=FALSE, message=F}

USArrests %>%
  mutate(State = row.names(.)) %>% # create var names state from rownames
  select(State, Murder) %>% # select vars desired from df
  filter(Murder > quantile(Murder, probs = .75)) %>% #filter to murders 75th+ percentile
  arrange(desc(Murder)) # sort by highest to lowest murders

```

# summarizing

## example of summarize

```{r summarize example, echo=TRUE}
USArrests %>%
 summarise(murder_mean = mean(Murder),
           murder_median = median(Murder)) 

```
## a word about group_by()
Most data operations are done by grouping. `group_by()` takes an existing table and converts it
into a grouped table where oprations are performed.

`ungroup()` removes grouping. 

## group_by + summarize powerful combo

```{r summarize example2, echo=TRUE}
USArrests$State <- row.names(USArrests) # add State var to df

USArrests %>%
  mutate(M_states = ifelse(grepl("^M", State), "Yes", "No")) %>% # identify if state starts w/ M
  select(-State) %>%
  group_by(M_states) %>% # summarize by M and non-M states
  summarise_all(list(mean = mean, median = median)) %>% # summarize all
  head()

```

# joining
 
## dplyr joins

**mutating joins** (combining variables from two data sources):

  - `inner_join()`
  - `left_join()`
  - `right_join()`
  - `full_join()`

**filtering joins** (keeps cases from the left table and use right to filter):

  - `semi_join()` retains only rows that both dfs have common AND only columns of left
  - `anti_join()` retains only rows that are non-existent in the right AND keeps only columns of left


## example of dplyr join in action

```{r join example, echo=TRUE, message=F}
USArrests %>%
  inner_join(cities, by = c("State" = "state_name")) %>% #join cities df by State 
  select(State, city, population) %>% # select vars
  sample_n(8) #return a sample of8

```

## example of anti-join
```{r join example2, echo=TRUE, message=F}

# show which states listed in cities df aren't found in USArrests
cities %>%
  anti_join(USArrests, by = c("state_name" = "State")) %>%
  distinct(state_name)
```

## nest_join() 

With the release of dplyr 0.8, a new join function was added called `nest_join()`.

`nest_join()` returns all rows and all columns from x. Adds a list column of tibbles, where
each tibble contains all the rows from y that match that row of x. 

```{r join example3, echo=TRUE, message=F}

nest_example <- cities %>%
  distinct(state_name) %>%
  filter(state_name %in% c('Kansas', 'Missouri')) %>%
  nest_join(USArrests, by = c("state_name" = "State")) 

# create a df from the returned nest_join list
bind_rows(nest_example$y) %>%
  mutate(state = nest_example$state_name) %>%
  select(state, everything()) # example of everything() usefule for column ordering
```

# misc tips

## what are the top 3 largest cities in ks & mo?

```{r misc example, echo=TRUE, message=F}
USArrests %>%
  select_if(is.character) %>% # selects State var since it's the only character var
  inner_join(select(.data = cities, state_name, city, population), # nested select
    by = c("State" = "state_name")) %>% # inner join 
  group_by(State) %>% # grouping state
  arrange(State, desc(population)) %>% # sorting
  filter(row_number() <= 3 & State %in% c('Kansas', 'Missouri')) # filter within group
```

## how to find unqiue rows?

```{r misc example2, echo=TRUE, message=F}
cities %>%
  select(state_name, timezone) %>%
  distinct() %>% # remove duplicate rows
  arrange(state_name) %>%
  head()
```

## selecting vars starting w/ 'state' & contain 'city'
```{r misc example3, echo=TRUE, message=F}
cities %>%
  select(starts_with("state"), contains("city")) %>%
  head()
```

## tally/count shortcuts
`tally()` is a convenient wrapper for summarise that will either call `n()` or `sum()`.`count()` is similiar but calls `group_by()`

```{r misc example5, echo=TRUE, message=F}
cities %>% tally() # shortcut for n
```


```{r misc example6, echo=TRUE, message=F}
cities %>% count(state_name) %>% head(3) # shortcut for group_by + tally
```


## case when example (similiar to SQL)
```{r misc example7, echo=TRUE, message=F}
USArrests %>%
  mutate(Murder_bin = case_when(Murder > median(.$Murder) ~ 'High',
                                Murder == median(.$Murder) ~ 'Middle',
                                TRUE ~ 'Low')) %>%
  count(Murder_bin) 

```

# questions?